Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-03-27 NovelLBM.tex
Update NovelLBM.tex

To illustrate our novel LBM algorithm, we consider two prototypical cell carrier fluids.
The first is an alginate solution which is often used in biofabrication but can also serve as a common food additive.
Our solution contains \SI{4}{\percent} alginate dissolved in PBS and was measured in a commercial plate-plate rheometer (see appendix~\ref{app:experiment} for experimental details).
The shear-dependent viscosity $\eta$ and the first normal stress difference $N_1$ are shown as function of shear rate in figure~\ref{fig:fit_alginate}.
We note the very high zero-shear viscosity of more than \SI{10000}{\milli\pascal\second}.
Bioinks in biofabrication are in general characterized by such high viscosities to guarantee stability of the printed construct.
At the same time, during the actual printing process when the material is flowing, the viscosity should be as low as possible to avoid excessive mechanical stress on the embedded cells.
This motivates the use of highly shear-thinning viscoelastic fluids in biofabrication.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_alginate_eta.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_alginate_N.pdf}
		\caption{Rheology data for alginate solution from 21 measurements of the viscosity (a) and normal stress (b).
				Orange lines show a fit of the PTT rheological model.}
		\label{fig:fit_alginate}
	\end{figure}

@@

The PTT model can be written in the form of a differential equation for the polymer stress $\ma{\tau}$ \cite{ferras2019}:
	\begin{equation}
		\overset{\nabla}{\ma{\tau}} = -\xi\left(\ma{\tau}\cdot\ma{D} + \ma{D}\cdot\ma{\tau}\right) - e^{\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}}\frac{\ma{\tau}}{\lambda} + 2\frac{\eta_\text{p}}{\lambda}\ma{D}
		\label{eq:polymerStress}
	\end{equation}
where $\ma{D}$ is the strain rate tensor.
We note that \cite{ferras2019} have also studied extended forms of the PTT model with a Mittag-Leffler instead of an exponential function, whereas here we restrict ourselves to the exponential version.
The polymer viscosity is denoted $\eta_\text{p}$ and $\lambda$ is the corresponding relaxation time.
$\epsilon$ is the so-called extensibility parameter which controls at which shear rates thinning takes place and $\xi$ describes the slip in the polymer network which we will set to zero in the following.
The upper convected derivative of an arbitrary tensor $\ma{A}$ is defined as follows:
	\begin{equation}
		\overset{\nabla}{\ma{A}} = \frac{\Dif\ma{A}}{\Dif t} - \left(\left(\nabla\vec{u}\right)^\mathrm{T}\cdot\ma{A} + \ma{A}\cdot\left(\nabla\vec{u}\right)\right)
	\end{equation}
where $\vec{u}$ is the fluid velocity and $\left(\nabla\vec{u}\right)_{ij} = \nabla_i\vec{u}_j$.
The material derivative is
	\begin{equation}
		\frac{\Dif\ma{A}}{\Dif t} = \frac{\partial\ma{A}}{\partial t} + \left(\vec{u}\cdot\nabla\right)\ma{A}.
	\end{equation}
\SG{ Kannst du der Vollständigkeit halber nochmal die Definition von $ \nabla \ma{ A }$ hinschreiben? }\RK{Nein}
Finally, the total fluid stress $\ma{\sigma}$ is the sum of the polymer stress $\ma{\tau}$ and a solvent stress due to the Newtonian solvent with viscosity $\eta_\text{s}$:
	\begin{align}
		\ma{\sigma} &= \ma{\tau} + 2\eta_\text{s}\ma{D} 		\label{eq:totalStress}
	\end{align}

For a simple shear flow in the $xy$-plane of a PTT fluid with $ \xi =0$, an analytical expression for $ \eta $ and $N_1$ can be derived as detailed in appendix~\ref{app:semianalyticalPoiseuille}:
	\begin{equation}
		\eta\left(\dot{\gamma}\right) = \frac{\eta_\text{p}}{\exp\left[0.5W_0\left(4\epsilon \mathrm{Wi}^2\right)\right]} + \eta_\text{s} \label{eq:shearEta}
	\end{equation}
	\begin{equation}
		N_1 = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon \mathrm{Wi}^2\right) \label{eq:shearN}
	\end{equation}
where $\mathrm{Wi} = \lambda\dot{\gamma}$ is the Weissenberg number and $W_0$ is the Lambert $W$-function.
Fitting these equations to the rheological data for alginate and methyl cellulose leads to satisfactory agreement as shown in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
In the fit, the solvent viscosity was fixed at $\eta_\text{s} = \SI{1}{\milli\pascal\second}$ as the viscosity of the PBS buffer. 
%\SG{ give ref }\RK{PBS is 99\% water. If you really want a citation tho we could use doi: 10.1063/1.4823586} \SG{ ok, let's leave it out. }
%At very high shear rates, the contained polymers are completely stretched and the viscosity asymptotically approaches the solvent value $ \eta _ \mathrm{s}$.
The obtained parameters are given in table~\ref{tab:PTTparameters}.
	\begin{table}[H]
		\centering
		\begin{tabular}{c|c|c|c|c}
			Fluid & $\eta_\text{p}/\unit{\milli\pascal\second}$ & $\lambda/\unit{\milli\second}$ & $\epsilon$ & $R_\eta$ \\
			\hline
			alginate \SI{4}{\percent} & \num{48.2\pm0.4e3} & \num{343 \pm 4} & \num{0.545 \pm 0.010} & \num{4.82e4} \\
			\rowcolor{secondaryBackground}
			methyl c. \SI{0.49}{\percent} & \num{18.7\pm0.4} & \num{0.344\pm0.003} & \num{0.270\pm0.003} & \num{18.7} \\
			methyl c. \SI{0.59}{\percent} & \num{32.5\pm0.7} & \num{0.433\pm0.004} & \num{0.365\pm0.004} & \num{32.5} \\
			\rowcolor{secondaryBackground}
			methyl c. \SI{0.83}{\percent} & \num{81\pm2} & \num{0.714\pm0.009} & \num{0.496\pm0.006} & \num{81}
		\end{tabular}
		\caption{Parameters of the PTT model obtained by fitting to the rheological measurements in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc} along with corresponding viscosity ratios from eq.~(\ref{eq:R_eta}).}
		\label{tab:PTTparameters}
	\end{table}

@@

Our next example is the RT-DC geometry in which cells from a reservoir are squeezed through a narrow observation channel \cite{Toepfner.2018, Fregin:2019gx} with a square cross-section of \qtyproduct{40x40}{\micro\meter}.
%$\SI{40}{\micro\meter} \times \SI{40}{\micro\meter}$.%Correctly typeset version, that compiles for Stephan.
For this simulation we use periodic boundary conditions and a constant body force corresponding to a pressure gradient of \SI{1e7}{\pascal\per\meter}. \SG{ units. Is this Pa/m or is this the pressure difference between inlet and outlet? }\RK{done}
In figure~\ref{fig:RT-DC}, we show the velocity profile, the von Mises stress and the shear component of the stress which is the relevant component if a local shear viscosity is to be defined.
There are three qualitative differences compared to a purely viscous fluid.
Directly after the inlet we find a region of low stress.
Due to the fast advection, the polymers, despite being inside the constriction, have not reacted to the increased fluid stress yet.
Similarly, directly after the constriction, the high stress inside the observation channel persists over a certain distance that is needed for the polymer stress to decay. If one would divide the stress in these two regions by the strain rate in an attempt to obtain a viscosity one would find a very low and a very high viscosity respectively. These can be outside the range of viscosities, that can be obtained through rheological measurements with the fluid. This effect is even more pronounced along the walls directly after the outlet. Here the viscosities obtained in this manner become negative. Due to the strain rate approaching $0$ at several locations within the simulation volume, this calculation often diverges. The clamped result can be seen in figure~\ref{fig:Nozzle/RT-DC_eta}. These discrepancies, especially the ones at the walls, show the breakdown of the viscosity interpretation in viscoelastic cases and signify the regions of interesting behavior. This also shows why viscoelastic simulations cannot always be replaced by purely viscous models and should be considered despite the high computational cost.

@@

The authors would like to thank the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) under Project number 326998133, TRR 225 Biofabrication (subprojects B07 and B09) and the Federal Ministry of Education and Research (BMBF) under Project SOP\_Bioprint under Contract Number 13XP5071B for enabling data collection.

@@

The solutions were prepared by dissolving alginate (DuPont VIVAPHARM Alginate PH176) in ultrapure water at a ratio of \SI{40}{\milli\gram\per\milli\liter}, which here is refereed to as \SI{4}{\percent}.
To ensure homogeneous dissolution, the material was stored at \SI{37}{\celsius} for \SI{24}{\hour} and regularly mixed to ensure complete dissolution.
The measurements were conducted with a \SI{25}{\milli\meter} plate-plate geometry at a \SI{500}{\micro\meter} gap on a rheometer (Anton paar MCR 702).
A solvent trap was used to limit evaporation.
All measurements were done at a set temperature of \SI{25}{\celsius} as shear rate sweep.
Data is agglomerated from 21 independent measurements.
The first normal stress difference $N_1$ is calculated from the measured normal force using the following equation \cite{kulicke1977}:
	\begin{equation}
		N_1 = \frac{2F}{\pi R^2} + \frac{3}{20}\rho\omega^2R^2
	\end{equation}
	where $F$ is the measured normal force, $R$ is the radius of the rheometer, $\rho$ is the density of the fluid and $\omega$ is the angular velocity of the rheometer. The second term is usually omitted and as we found it to matter little, and as the quality of the data we have available is too limited to warrant such detailed corrections, we also choose to omit it. A cone-rheometer would be preferred here, but as $N_2 = 0$ for PTT with $\xi = 0$, the difference is negligible.

@@

As additional quantitative validation, we simulate a pure shear flow driven by imposing a fixed velocity at the top and bottom of the computational box and using periodic boundary conditions in the other two directions.
The box dimension is 2x43x2.
We use PTT with the methyl cellulose parameters and vary the shear rate to cover the full range starting from the zero-shear all the way to the infinite-shear plateau displayed in figure~\ref{fig:fit_mc} of the main text.
The shuffling is fixed at $\alpha_\text{s} = 1$ and the viscosity as function of shear rate is computed by dividing the total fluid stress at the center of the box by the shear rate.
Figure~\ref{fig:wiParameterStudy}~(a) shows that our simulations are in very good agreement with the semi-analytical solution with a maximum error of \num{2e-11}.
Importantly, we show in figure~\ref{fig:wiParameterStudy}~(b) that this very good agreement is independent of the shuffling parameter $\alpha_\text{s}$.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/wiParameterStudy.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/suffleParameterStudy.pdf}
		\caption{(a) The local viscosity determined by a shear flow simulation is in very good agreement with the semi-analytical theory.
		(b) The error is negligibly small and independent of the shuffling parameter $\alpha_\text{s}$. \SG{ at which $ \dot{ \gamma  }$? }\RK{\SI{1}{\per\second}}}
		\label{fig:wiParameterStudy}
	\end{figure}

>>>>

To illustrate our novel LBM algorithm, we consider two prototypical cell carrier fluids.
The first is an alginate solution which is often used in biofabrication but can also serve as a common food additive.
Our solution contains \SI{4}{\percent} w/v alginate dissolved in ultrapure water and was measured in a commercial plate-plate rheometer (see appendix~\ref{app:experiment} for experimental details).
The shear-dependent viscosity $\eta$ and the first normal stress difference $N_1$ are shown as function of shear rate in figure~\ref{fig:fit_alginate}.
We note the very high zero-shear viscosity of more than \SI{10000}{\milli\pascal\second}.
Bioinks in biofabrication are in general characterized by such high viscosities to guarantee stability of the printed construct.
At the same time, during the actual printing process when the material is flowing, the viscosity should be as low as possible to avoid excessive mechanical stress on the embedded cells.
This motivates the use of highly shear-thinning viscoelastic fluids in biofabrication.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_alginate_eta.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/fit_alginate_N.pdf}
		\caption{Rheology data for alginate solution from 21 measurements of the viscosity (a) and normal stress (b).
				Orange lines show a fit of the PTT rheological model.}
		\label{fig:fit_alginate}
	\end{figure}

@@

The PTT model can be written in the form of a differential equation for the polymer stress $\ma{\tau}$ \cite{ferras2019}:
	\begin{equation}
		\overset{\nabla}{\ma{\tau}} = -\xi\left(\ma{\tau}\cdot\ma{D} + \ma{D}\cdot\ma{\tau}\right) - e^{\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}}\frac{\ma{\tau}}{\lambda} + 2\frac{\eta_\text{p}}{\lambda}\ma{D}
		\label{eq:polymerStress}
	\end{equation}
where $\ma{D}$ is the strain rate tensor.
We note that \cite{ferras2019} have also studied extended forms of the PTT model with a Mittag-Leffler instead of an exponential function, whereas here we restrict ourselves to the exponential version.
The polymer viscosity is denoted $\eta_\text{p}$ and $\lambda$ is the corresponding relaxation time.
$\epsilon$ is the so-called extensibility parameter which controls at which shear rates thinning takes place and $\xi$ describes the slip in the polymer network which we will set to zero in the following.
The upper convected derivative of an arbitrary tensor $\ma{A}$ is defined as follows:
	\begin{equation}
		\overset{\nabla}{\ma{A}} = \frac{\Dif\ma{A}}{\Dif t} - \left(\left(\nabla\vec{u}\right)^\mathrm{T}\cdot\ma{A} + \ma{A}\cdot\left(\nabla\vec{u}\right)\right)
	\end{equation}
where $\vec{u}$ is the fluid velocity and $\left(\nabla\vec{u}\right)_{ij} = \nabla_i\vec{u}_j$.
The material derivative is
	\begin{equation}
		\frac{\Dif\ma{A}}{\Dif t} = \frac{\partial\ma{A}}{\partial t} + \left(\vec{u}\cdot\nabla\right)\ma{A}.
	\end{equation}
Finally, the total fluid stress $\ma{\sigma}$ is the sum of the polymer stress $\ma{\tau}$ and a solvent stress due to the Newtonian solvent with viscosity $\eta_\text{s}$:
	\begin{align}
		\ma{\sigma} &= \ma{\tau} + 2\eta_\text{s}\ma{D} 		\label{eq:totalStress}
	\end{align}

For a simple shear flow in the $xy$-plane of a PTT fluid with $ \xi =0$, an analytical expression for $ \eta $ and $N_1$ can be derived as detailed in appendix~\ref{app:semianalyticalPoiseuille}:
	\begin{equation}
		\eta\left(\dot{\gamma}\right) = \frac{\eta_\text{p}}{\exp\left[0.5W_0\left(4\epsilon \mathrm{Wi}^2\right)\right]} + \eta_\text{s} \label{eq:shearEta}
	\end{equation}
	\begin{equation}
		N_1 = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon \mathrm{Wi}^2\right) \label{eq:shearN}
	\end{equation}
where $\mathrm{Wi} = \lambda\dot{\gamma}$ is the Weissenberg number and $W_0$ is the Lambert $W$-function.
Fitting these equations to the rheological data for alginate and methyl cellulose leads to satisfactory agreement as shown in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc}.
In the fit, the solvent viscosity was fixed at $\eta_\text{s} = \SI{1}{\milli\pascal\second}$ as the viscosity of the PBS buffer. 
%\SG{ give ref }\RK{PBS is 99\% water. If you really want a citation tho we could use doi: 10.1063/1.4823586} \SG{ ok, let's leave it out. }
%At very high shear rates, the contained polymers are completely stretched and the viscosity asymptotically approaches the solvent value $ \eta _ \mathrm{s}$.
The obtained parameters are given in table~\ref{tab:PTTparameters}.
	\begin{table}[H]
		\centering
		\begin{tabular}{c|c|c|c|c}
			Fluid & $\eta_\text{p}/\unit{\milli\pascal\second}$ & $\lambda/\unit{\milli\second}$ & $\epsilon$ & $R_\eta$ \\
			\hline
			alginate \SI{4}{\percent} w/v & \num{48.2\pm0.4e3} & \num{343 \pm 4} & \num{0.545 \pm 0.010} & \num{4.82e4} \\
			\rowcolor{secondaryBackground}
			methyl c. \SI{0.49}{\percent} & \num{18.7\pm0.4} & \num{0.344\pm0.003} & \num{0.270\pm0.003} & \num{18.7} \\
			methyl c. \SI{0.59}{\percent} & \num{32.5\pm0.7} & \num{0.433\pm0.004} & \num{0.365\pm0.004} & \num{32.5} \\
			\rowcolor{secondaryBackground}
			methyl c. \SI{0.83}{\percent} & \num{81\pm2} & \num{0.714\pm0.009} & \num{0.496\pm0.006} & \num{81}
		\end{tabular}
		\caption{Parameters of the PTT model obtained by fitting to the rheological measurements in figures~\ref{fig:fit_alginate} and \ref{fig:fit_mc} along with corresponding viscosity ratios from eq.~(\ref{eq:R_eta}).}
		\label{tab:PTTparameters}
	\end{table}

@@

Our next example is the RT-DC geometry in which cells from a reservoir are squeezed through a narrow observation channel \cite{Toepfner.2018, Fregin:2019gx} with a square cross-section of \qtyproduct{40x40}{\micro\meter}.
%$\SI{40}{\micro\meter} \times \SI{40}{\micro\meter}$.%Correctly typeset version, that compiles for Stephan.
For this simulation we use periodic boundary conditions and a constant body force corresponding to a pressure gradient of \SI{1e7}{\pascal\per\meter}.
In figure~\ref{fig:RT-DC}, we show the velocity profile, the von Mises stress and the shear component of the stress which is the relevant component if a local shear viscosity is to be defined.
There are three qualitative differences compared to a purely viscous fluid.
Directly after the inlet we find a region of low stress.
Due to the fast advection, the polymers, despite being inside the constriction, have not reacted to the increased fluid stress yet.
Similarly, directly after the constriction, the high stress inside the observation channel persists over a certain distance that is needed for the polymer stress to decay. If one would divide the stress in these two regions by the strain rate in an attempt to obtain a viscosity one would find a very low and a very high viscosity respectively. These can be outside the range of viscosities, that can be obtained through rheological measurements with the fluid. This effect is even more pronounced along the walls directly after the outlet. Here the viscosities obtained in this manner become negative. Due to the strain rate approaching $0$ at several locations within the simulation volume, this calculation often diverges. The clamped result can be seen in figure~\ref{fig:Nozzle/RT-DC_eta}. These discrepancies, especially the ones at the walls, show the breakdown of the viscosity interpretation in viscoelastic cases and signify the regions of interesting behavior. This also shows why viscoelastic simulations cannot always be replaced by purely viscous models and should be considered despite the high computational cost.

@@

The authors would like to thank the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) under Project number 326998133, TRR 225 Biofabrication (subprojects B07, B09 and Z01) and the Federal Ministry of Education and Research (BMBF) under Project SOP\_Bioprint (Contract Number 13XP5071B) for enabling data collection.

@@

The solutions were prepared by dissolving alginate (DuPont VIVAPHARM Alginate PH176) in ultrapure water at a ratio of \SI{40}{\milli\gram\per\milli\liter}, which here is refereed to as \SI{4}{\percent} w/v.
To ensure homogeneous dissolution, the material was stored at \SI{37}{\celsius} for \SI{24}{\hour} and regularly mixed to ensure complete dissolution.
The measurements were conducted with a \SI{25}{\milli\meter} plate-plate geometry at a \SI{500}{\micro\meter} gap on a rheometer (Anton paar MCR 702).
A solvent trap was used to limit evaporation.
All measurements were done at a set temperature of \SI{25}{\celsius} as shear rate sweep.
Data is agglomerated from 21 independent measurements.
The first normal stress difference $N_1$ is calculated from the measured normal force using the following equation \cite{kulicke1977}:
	\begin{equation}
		N_1 = \frac{2F}{\pi R^2} + \frac{3}{20}\rho\omega^2R^2
	\end{equation}
	where $F$ is the measured normal force, $R$ is the radius of the rheometer, $\rho$ is the density of the fluid and $\omega$ is the angular velocity of the rheometer. The second term is usually omitted and as we found it to matter little, and as the quality of the data we have available is too limited to warrant such detailed corrections, we also choose to omit it. A cone-rheometer would be preferred here, but as $N_2 = 0$ for PTT with $\xi = 0$, the difference is negligible.

@@

As additional quantitative validation, we simulate a pure shear flow driven by imposing a fixed velocity at the top and bottom of the computational box and using periodic boundary conditions in the other two directions.
The box dimension is 2x43x2.
We use PTT with the methyl cellulose parameters and vary the shear rate to cover the full range starting from the zero-shear all the way to the infinite-shear plateau displayed in figure~\ref{fig:fit_mc} of the main text.
The shuffling is fixed at $\alpha_\text{s} = 1$ and the viscosity as function of shear rate is computed by dividing the total fluid stress at the center of the box by the shear rate.
Figure~\ref{fig:wiParameterStudy}~(a) shows that our simulations are in very good agreement with the semi-analytical solution with a maximum error of \num{2e-11}.
Importantly, we show in figure~\ref{fig:wiParameterStudy}~(b) that this very good agreement is independent of the shuffling parameter $\alpha_\text{s}$.
	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/wiParameterStudy.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/suffleParameterStudy.pdf}
		\caption{(a) The local viscosity determined by a shear flow simulation is in very good agreement with the semi-analytical theory.
		(b) The error is negligibly small and independent of the shuffling parameter $\alpha_\text{s}$. Shown at \SI{1}{\per\second} and .}
		\label{fig:wiParameterStudy}
	\end{figure}\todo{1e5; maybe write this in legend}
